Color- and piece-blind chess

### 2019-03-08 chess/papers/blind.tex
More hacking on sidetrack, need to rewrite again

Playing multiple games simutaneously is an easy extension of playing a
single game, although in principle the scale of such a thing could
still be impressive. With a modern chess engine, it is easy to scale
to an arbitrary number of games, since the exhibitor can make progress
by observing one of the boards, computing a strong move, and playing
it; this requires $O(0)$ space because all of the state is stored
externally in the exhibition itself. 
% If playing blindfold, then the
% main cost is storing the state of each game. 
However, we run the risk of losing the tournament (other players may
be yet stronger computers). The man-in-the-middle attack remains an
efficient way way to minimize loss (ensuring an exactly even score).
Suppose that there are $n$ games $G$ being played. We generate a
bipartite matching that pairs games $\langle G_w, G_b \rangle$ (where
we play as white against Bob and black against Walice, respectively).
This is $n/2$ pairs, and we record which pair each game belongs to, in
$n \log_2(n/2)$ bits. Now, to make progress, we look at a game $G_i$
where it is our move, and consult our state for this pair. If the
pair has the has-move bit (actually just store it as the move itself),

ok, so probably the most efficient thing is to store instead the
move played in each row. Moves are like 64*64*4 (promotion) = 6+6+2 bits.
The only way to really motivate the board approach would be if it
allows us to avoid the log(n/2) factor from storing the matching?

@@

I think a version I liked was 64 bits (this gives the positions of all
pieces), followed by 32 records giving the piece values (there are 6:
pawn, knight, bishop, rook, queen, king) and colors. To pick between
the 12 possibilities we need 4 bits, and there is some slack here of
course. anyway, that's 64 + 32*4 = 192 bits, plus 15 bits for other
state. Note we could use the c\_rook trick to save 4 bits for castling
rights (14 possibilities), not to mention a en\_passantable pawn (15)
and a king whose turn it is (even 16). Clear slack here, but that
would leave only the 50-move rule at 7 bits, for a total of 199.
Wikipedia says the huffman approach is ``maximum of 204 bits, and
often much less.'' but the article also contains many bugs, like
the misconception that there can only be four rooks.

>>>>

Playing multiple games simutaneously is an easy extension of playing a
single game, although in principle the scale of such a thing could
still be impressive. With a modern chess engine, it is easy to scale
to an arbitrary number of games, since the exhibitor can make progress
by observing one of the boards, computing a strong move, and playing
it; this requires $O(0)$ space because all of the state is stored
externally in the exhibition itself. 
% If playing blindfold, then the
% main cost is storing the state of each game. 
However, we run the risk of losing the tournament (other players may
be yet stronger computers). The man-in-the-middle attack remains an
efficient way way to minimize loss (ensuring an exactly even score).
The simplest way to do this is to explicily generate a perfect
bipartite matching over the $n$ games $G$ being played. This consists
of $n/2$ pairs $\langle G_w, G_b \rangle$ (where we play as white
against Bob and black against Walice, respectively). Since each game
starts in the starting position, this is very easy; we can just assign
the matches consecutively. Along with each pair we also record which
of the following states we are in:
\begin{enumerate}
\item We are waiting for a move from Walice (our white opponent) \label{state:mim1}
\item We have seen a move from Walice, which is \underline{\hspace{6em}}. \label{state:mim2}
\item We are waiting for a move from Bob (our black opponent) \label{state:mim3}
\item We have seen a move from Bob, which is \underline{\hspace{6em}}. \label{state:mim4}
\end{enumerate}
If in State~\ref{state:mim1}, we just watch $G_b$ until Walice makes a
move, then record it and proceed to State~\ref{state:mim2}. We consume
the move and move to State~\ref{state:mim3} by playing that move in
$G_w$ against Bob (where it must be our turn). We can immediately seek
out that game or wait until we naturally come upon it. However, we
should only approach $G_w$ when the pair of games is in
State~\ref{state:mim3}, etc., otherwise we will not have a move to play.

There are $n/2$ pairs, with two bits for the state, no more\footnote{
  There are only 1792 pairs of squares between which pieces can ever
  move (Section~\ref{sec:legalmove}), so $11+2$ bits suffices, with
  some added complexity.} than $\log_2(64 * 64 * 4) = 14$ bits for
each move (source square, destination square, and 2 bits to
distinguish promotion to queen, rook, bishop, or knight). However, we
also need to store the matching of $G_w$ to $G_b$; this can be done
with a pair of indices (or e.g. memory addresses) but unfortunately,
this requires $\log_2(n)$ bits to represent. So overall this approach
requires $O(n \log(n))$ space to play a simultaneous exhibition of $n$
games.

It appears to be possible to reduce the space usage per game to a
constant. In order to perform a man-in-the-middle attack, we need
a perfect matching between the white games and black games. It is
not essential that the matching be stable over time; for example
if we are forwarding moves between Walice and Bob, and between
Waluigi and Bario, and these games happen to transpose to the same
position, then it works just fine to switch to forwarding between
Walice and Bario; Waluigi and Bob. So, rather than store the
matching explicitly, we can reconstruct it from the stored state.
We two vectors of $n/2$ rows, the ``source'' and ``destination'' rows,  (XXX I think we do not need destination)
each containing a game {\em position} (not move).
This position represents XXXX (like the last time we saw it?).
The positions are initialized to the starting position. Like before
we can be in one of four states:
\begin{itemize}
\item Waiting for all ``source'' players (who play as white) to complete their moves. \label{state:cmim1}
\item Forwarding our moves from ``source'' to our black opponents. \label{state:cmim2}
\item `` \ldots '' (who play as black) `` \ldots '' \label{state:cmim3}
\item `` \ldots '' white opponents. \label{state:cmim3}
\end{itemize}

\newcommand\stepsto[1]{\stackrel{#1}{\rightarrow}}

% (XXX So actually, simpler here is just: Loop until all white players have
% made their moves. Then copy all their boards to our board vector.)
While in State~\label{state:cmim1} (which we also start in), we
compare our stored source board states to the boards where we play as
black. Once the white opponent has made a move, resulting in $B'$,
there will be some position $B$ in our ``source'' set that is a valid
predecessor for $B'$ (that is, there exists some legal move $M$ that
advances $B$ to $B'$). For example, at the beginning, $B$ may be the
starting position (in fact it must, since all boards are initialized
this way) and $B'$ may be the position resulting from $1. e4$. With
the matching constructed (see below) we copy each $B'$ into our set of
boards, forgetting $B$.\footnote{In fact, we do not even need to
  construct the matching, although it may help us with bookkeeping
  (e.g., which games have already been copied?), depending on how the
  exhibition itself is represented.} Then we transition to
State~\label{state:cmim2}. In this state, we figure out what moves to
play in the pending boards. By invariant, each one $B$ will be
compatible with at least one of our source boards $B'$. Compatible
means that there exists a move $m$ where playing $m$ in board $B$
results in $B'$. If $m$ exists, it is unique (and easy to compute), so
this is the move that we will play. The only tricky thing is figuring
out which boards go with which; this is the problem of reconstructing
the perfect matching.

Although if $B \stepsto{m} B'$ it is easy to deduce $m$, it is not
possible to compute $B$ from $B'$, or even from $B'$ and $m$. This is
because we may have both $B_a \stepsto{m_a} B'$ and $B_b \stepsto{m_b}
B'$ with $B_a \neq B_b$. For example with $B'$

we could have $B_a$ and $B_b$ be

\chessboard[tinyboard, setfen=Nk6/8/8/8/8/8/8/K7 b - -, showmover]
\chessboard[tinyboard, setfen=Rk6/8/8/8/8/8/8/K7 b - -, showmover]

Both of which can precede $B'$ (the move is even the same: Kxa1).
So it is not enough to greedily assign edges in our perfect match;
if we choose the edge $B_b$ to go with $B'$, we might later find
$B'_2$:

\ldots and have no possible matching board, since it cannot follow
$B_a$.

Fortunately, we know the
 - there exists a perfect matching (players playing legally)
 - we can tell if we found one

@@

I think a version I liked was 64 bits (this gives the positions of all
pieces),\footnote{(We can also do a little better. Note that only up
  to 32 of the 64 bits can be set, so you could represent
  $\binom{60}{32}$ in $\sim 61$ bits and use e.g. a third king to
  indicate that there are no more actually occupied squares if fewer
  than 32 pieces are present. However, this gets vastly more
  complicated for only 3 bits of savings.)} followed by 32 records
giving the piece values (there are 6: pawn, knight, bishop, rook,
queen, king) and colors. To pick between the 12 possibilities we need
4 bits, and there is some slack here of course. anyway, that's 64 +
32*4 = 192 bits, plus 15 bits for other state. Note we could use the
c\_rook trick to save 4 bits for castling rights (14 possibilities),
not to mention a en\_passantable pawn (15) and a king whose turn it is
(even 16). Clear slack here, but that would leave only the 50-move
rule at 7 bits, for a total of 199. Wikipedia says the huffman
approach is ``maximum of 204 bits, and often much less.'' but the
article also contains many bugs, like the misconception that there can
only be four rooks.
